Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-24 paper/paper.tex
Add todos
pass through mechanization section

@@

\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{tabular}{ c |  c  | c  }
    & Consistency & Normalization \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & - \\
    Renaming & 46 & -  \\
    % Syntactic soundness & 660 &  - \\
    Untyped reduction & 349 & - \\
    Neutral and normal forms & N/A & 269 \\
    Logical relation & 331 & 426 \\
    Semantic typing and soundness & 169 & 199 \\
    \hline
    Total & 963 & 1357 \\
  \end{tabular}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

Figure~\ref{fig:linecount} shows the statics of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the
extension to normalization for open and closed terms from
Section~\ref{sec:extension}.

Since the normalization and the consistency development share the same
syntactic and typing specification, they differ in only the categories
that are related to the logical relation. We use - as a marker that
the line count is the same.

@@

Orthogonal to the development of our consistency and normalization
proof, we prove the syntactic soundness of our system through subject
reduction and progress. The syntactic soundness proof shares the
definition of renaming with our semantic soundness proof, which is
factored out as a separate file under the Renaming (common) category.

\subsection{Artifacts Specific to Coq}
In this section, we discuss the Coq encoding of the definitions and proofs presented
in Section~\ref{sec:logrelproof}  and the artifacts that are
specific to Coq and may not appear in other proof assistants.

@@

In our proof presented in Section~\ref{sec:logreldep}, the relation
$[[R]]$ in \rref{I-Pi} requires Coq's impredicative \texttt{Prop} sort
since $[[R]]$ can later be instantiated into something defined in
terms of the logical relation itself (e.g. in
Lemma~\ref{lemma:piintroalt}). However, in the mechanized proof, the
definition of $[[Interp I i A S]]$ can in fact be encoded without the
use of impredicativity by consistently replacing the use of
\texttt{Prop} with Coq's predicative sort \texttt{Type}. One can even
use this \texttt{Type} variant of the logical relation to define
interpretation for any finite number of universes. The lack of
impredicativity becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation defined in
Figure~\ref{fig:logrelrec} that recursively calls itself at smaller
universe levels. As a result, the one feature that truly requires
impredicativity from \lang{} is its countable universe
hierarchy, though lemmas such as the admissibility of \rref{I-PiAlt}
(Lemma~\ref{lemma:piintroalt}) not only simply our mechanization, but
also give better intuition on what our logical relation means.

The definition of $[[Interp I i A S]]$ has an almost one-to-one
correspondence to the Coq definition. The main difference is the
specification of $[[I]]$. In the presentation in
Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
numbers less than $[[i]]$, the universe level. In Coq, we only require
$[[I]]$ to be function with the set of natural numbers as its domain.
In the Coq encoding of $[[InterpR i A S]]$ (Figure~\ref{fig:logrelrec}), we define $[[I]]$ as follows.
\begin{equation*}
  \begin{split}
    [[I]]   &\in [[SNat -> PowerSet STm]] \\
    [[I j]] &= [[{A | exists S , InterpR j A S}]] \text{, if } j < i \\
    [[I j]] &= [[emptyset]] \text{, } j \geq i
  \end{split}
\end{equation*}
Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
\rref{I-Set}, we can retroactively show that the set we return in the $j
\geq i$ case is junk data that does not affect the result of the logical
relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Figure~\ref{fig:logrelrec}.

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
The triangle property states
that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
is the Takahashi translation~\citep{takahashi-parallel-reduction}
which roughly corresponds to simultaneous reduction of the redexes in
$[[a]]$, excluding the new redexes that appear as a result of
reduction.
\begin{minted}{coq}
Lemma par_triangle a : forall b, Par a b ->  Par b (tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the
graph of \mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par} and
\texttt{ctrs:Par} say that the proof involves inverting and
constructing of the derivations of parallel
reduction. The option \texttt{use:Par\_refl,par\_cong} allows the
automation tactic to use
the reflexivity and congruence properties of parallel
reduction as lemmas.
The flag \texttt{lq:on} toggles tunes CoqHammer's search
algorithm, though we never need to specify those options
manually. Instead, we first invoke the \texttt{sauto} tactic indirectly through the
\texttt{best} tactic provided by CoqHammer where we specify only the
\texttt{inv}, \texttt{ctrs}, and lemmas we want to use. The
\texttt{best} tactic iterates through possible configurations and
provides us a replacement to the \texttt{best} tactic with the
tuned performance flags to save time for future re-execution of the
tactic.

The automation provided by CoqHammer not only gives us a proof that is
shorter and more resiliant to changes, but also gives useful
documentation for readers who wish to understand how the underlying
proof works since the automation tactic is guaranteed to not use
lemmas or invert derivations that are not specified in the
\texttt{use} or \texttt{inv} flag.

>>>>

%% to allow unicode characters in minted code
\usepackage{newunicodechar}
\newunicodechar{⇒}{$\Rightarrow$}

@@

\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{tabular}{ l |  c  | c | c }
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    % Library (Autosubst 2)  & 491 & - \\
    Syntactic typing (specification) &  68 & - & - \\
    Renaming & 46 & -  & - \\
    Untyped reduction & 349 & - & - \\
    Neutral and normal forms & N/A & 269 & N/A \\
    Logical relation & 331 & 426 & N/A \\
    Semantic typing and soundness & 169 & 199 & N/A \\
    Syntactic soundness & N/A & N/A  & 660 \\
    \hline
    Total & 963 & 1357 & 1123 \\
  \end{tabular}
  \caption{Statistics of the Coq Development}
  \label{fig:linecount}
\end{figure}

Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool} for each file of our
development, including the base consistency proof from
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} and the extension to
normalization for open and closed terms from Section~\ref{sec:extension}.

Because the normalization and the consistency development share the same
syntactic and typing specification, they differ in only the categories
that are relevant to the logical relation. We use - as a marker that
the line count is the same.

@@

Orthogonal to the development of our consistency and normalization
proof, we have also proven the syntactic soundness of our system through subject
reduction and progress. The syntactic soundness proof shares the
definition of renaming with our semantic soundness proof, which is
factored out as a separate file under the Renaming (common) category.
\scw{Should add this line count as another column, for comparison. I've done so, 
but may need to update the numbers. }

\subsection{Artifacts Specific to Coq}
In this section, we discuss the Coq encoding of the definitions and proofs presented
in Section~\ref{sec:logrelproof} and the artifacts of this encoding that are
specific to Coq.

@@

In our proof presented in Section~\ref{sec:logreldep}, the relation
$[[R]]$ in \rref{I-Pi} requires Coq's impredicative \texttt{Prop} sort
since $[[R]]$ can later be instantiated into something defined in
terms of the logical relation itself (e.g. in
Lemma~\ref{lemma:piintroalt}). \scw{I'm having trouble connecting the rule in 
the figure with rule \texttt{InterpExtFun} in the coq development. Would it make 
sense to include the definition of ProdSpace? What is the Coq type of R?}

In the Coq mechanized proof, the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, when \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S}
has type \texttt {tm -> Prop}.  However, if desired, we could consistently
replace the use of \texttt{Prop} with Coq's predicative sort
\texttt{Type}. This variation of the definition could be used to define the
interpretation for any \emph{finite} number of universes. The use of
\texttt{Type} becomes troublesome only when we attempt to define
$[[InterpR i A S]]$, the top-level logical relation defined in
Figure~\ref{fig:logrelrec} that recursively calls itself at smaller universe
levels. Therefore, the one feature of \lang{} that truly requires impredicativity 
is its countable universe hierarchy. \scw{I'm not sure the last part of this sentence
adds much: ``Although, lemmas such as the
admissibility of \rref{I-PiAlt} (Lemma~\ref{lemma:piintroalt}) not only simply
our mechanization, but also give better intuition on what our logical relation
means.'' Yes, we need impredicativity to define these lemmas, but without impredicativity
our logical relation means something else.}

The definition of $[[Interp I i A S]]$ has an almost one-to-one
correspondence to the Coq definition. The main difference is the
specification of $[[I]]$. In 
Section~\ref{sec:logreldep}, we define $[[I]]$ as a function over
numbers less than $[[i]]$, the universe level. In Coq, we only require
$[[I]]$ to be function with the set of natural numbers as its domain.
In the Coq encoding of $[[InterpR i A S]]$ (Figure~\ref{fig:logrelrec}), we define $[[I]]$ as follows.
\begin{equation*}
  \begin{split}
    [[I]]   &\in [[SNat -> PowerSet STm]] \\
    [[I j]] &= 
     \begin{cases}
      \ [[{A | exists S , InterpR j A S}]] & \text{when } j < i \\
      \ [[emptyset]] & \text{otherwise} 
    \end{cases}
  \end{split}
\end{equation*}
Since $[[I]]$ is only applied to numbers strictly less than $[[i]]$ in
\rref{I-Set}, we can retroactively show that the set we return in the $j
\geq i$ case is junk data that does not affect the result of the logical
relation. This property allows us to recover the simple equation for $[[InterpR i A S]]$ shown in Figure~\ref{fig:logrelrec}.

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
The triangle property states
that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
is the Takahashi translation~\citep{takahashi-parallel-reduction}
which roughly corresponds to simultaneous reduction of the redexes in
$[[a]]$, excluding the new redexes that appear as a result of
reduction.
\begin{minted}{coq}
Lemma par_triangle a : forall b, (a ⇒ b) -> (b ⇒ tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the graph of
\mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par}
and \texttt{ctrs:Par} say that the proof involves inverting and constructing
of the derivations of parallel reduction. The option
\texttt{use:Par\_refl,par\_cong} allows the automation tactic to use the
reflexivity and congruence properties of parallel reduction as lemmas.

The flag \texttt{lq:on} tunes CoqHammer's search algorithm.  While this flag
appears arcane, when developing our proof scripts we never specify this option
manually. Instead, we first invoke the \texttt{best} tactic provided by
CoqHammer, specifying only the \texttt{inv}, \texttt{ctrs}, and lemmas that we
want to use. The \texttt{best} tactic then iterates through possible
configurations and provides us with a replacement with the tuned performance
flags that save time for future re-execution of the proof script.

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resiliant to changes, but also gives useful documentation for readers
who wish to understand how the underlying proof. Although automation performs
extensive search, it does not use lemmas or invert derivations that are not
specified in the \texttt{use} or \texttt{inv} flag.

\subsection{Use of impredicativity}
\yl{Mention that logrel itself doesn't require impredicativty}
\scw{Didn't we say this above?}

\subsection{Extraction of conversion algorithm}
\yl{technically future work. Not sure whereelse to put it}
